---
title: "Ashotsk"
editor: visual
execute: 
  echo: false
  warning: false
---

```{r message=FALSE}
library(tidyverse)
ashotsk <- read_csv("ashotsk.csv")
```

Ashotsk is a village in the Ashotsk Municipality of the Shirak Province of Armenia.

### Fire causes and risk
Now we will explore the fire causes probability in Ashotsk, going through data provided by Tumo Climate Net. The main climate features that can cause fire are temperature and humidity. The fire cause probability is higher when the temperature is high and humidity is low. The graph below represents the temperature and humidity flow during June-July 2024. Dangerous humidity levels were registered from the end of June to the start of July.

```{r message=FALSE}
library(ggplot2)
library(dplyr)
library(patchwork)
p1 <- ashotsk |> drop_na() |> ggplot(aes(x=time, y=temperature)) + geom_line(color="darkblue") + geom_smooth(color="white") + theme_minimal() + labs(subtitle="Dangerous humidity levels were registered from the end of June to the start of July", x=NULL, y="Temperatue", title="Temperature and Humidity in Ashotsk") + scale_x_datetime(
    limits = c(as.POSIXct("2024-06-20 00:00"), as.POSIXct("2024-07-15 00:00")),
    date_labels = "%b %d"
  )

p2 <- ashotsk |> drop_na() |> ggplot(aes(x=time, y=humidity)) + geom_line(aes(color=if_else(humidity<30,"too_low", "normal")))+ scale_color_manual(values = c("too_low"="red", normal="orange"), guide="none") + theme_minimal() + labs(x="Time", y="Humidity") + scale_x_datetime(
    limits = c(as.POSIXct("2024-06-20 00:00"), as.POSIXct("2024-07-15 00:00")),
    date_labels = "%b %d"
  )

combined_plot <- p1/p2

print(combined_plot)
```

Humidity level is considered to be low when lower than 30 percents. The graph below represents the humidity level in Ashotsk when the temperature is above 30 C (June-July 2024).

```{r message=FALSE}
library(ggplot2)
library(dplyr)
library(plotly)
library(lubridate)

p <- ashotsk |> drop_na() |> filter(temperature > 30) |> mutate(humidity_level = factor(if_else(humidity < 30, "Too low", "Normal"))) |> ggplot(aes(x = time, y = temperature, alpha = 0.5)) + geom_point(aes(color = humidity_level)) + scale_color_manual(values = c("Too low" = "firebrick1", "Normal" = "steelblue1")) + theme_bw() + labs(title="Humidity level in Ashotsk", subtitle="Humidity levels are too low when temperatue is higher than 30 C", color = "Humidity level", x = "Time", y = "Temperature") + scale_alpha_continuous(guide = "none") + scale_x_datetime(
    limits = c(as.POSIXct("2024-06-20 00:00"), as.POSIXct("2024-07-15 00:00")),
    date_labels = "%b %d"
  )

ggplotly(p)

```

### Predicting the spread of fire

We never know when and where will start the fire, so predicting the fire spreading speed and direction would be helpful to put it out. In the graph below you can see the wind speed during June-July 2024. There is not much danger.

```{r message=FALSE}
ashotsk |> drop_na() |> ggplot(aes(x=time, y=speed)) + geom_jitter(aes(color=if_else(speed<15,"normal", "too_high"), alpha=0.5))+ scale_color_manual(values = c("normal"="black", too_high="darkred"), labels=c("normal"="Normal Speed", "too_high"="High Speed")) + scale_alpha_continuous(guide = "none") + geom_smooth(color="white") + theme_minimal() + labs(title="Wind speed in Ashotsk", x ="Time", y="Speed", color="Wind speed", subtitle="The wind speed is usually normal") + scale_x_datetime(
    limits = c(as.POSIXct("2024-06-20 00:00"), as.POSIXct("2024-07-15 00:00")),
    date_labels = "%b %d"
  )
```

It is important to predict the direction of fire spread as well. Here is a chart representing the count of each of wind direction. As you can see, the wind often blows from West.

```{r message=FALSE}
order <- c("NE", "E", "SE", "S", "SW", "W", "NW", "N")

p_circular <- ashotsk |> drop_na() |> mutate(direction = factor(direction, levels = order)) |>
  ggplot(aes(x=as.factor(direction))) + 
  geom_bar(fill="skyblue4") + coord_polar(start = 0) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.title.x = element_blank(),  
    axis.title.y = element_blank(), 
    panel.grid = element_blank(),
    plot.background = element_blank()
  ) + labs(title="Wind direction counts in Ashotsk", subtitle="The wind often blows from West")

print(p_circular) 

start_date <- as.POSIXct('2024-06-20 00:00')
end_date <- as.POSIXct('2024-07-15 00:00')

filtered_data <- ashotsk |>
  filter(time >= start_date & time <= end_date)

column_values <- filtered_data[["humidity"]]
column_values_numeric <- as.numeric(column_values)
mean_value <- mean(column_values_numeric, na.rm = TRUE)
#print(mean_value)
```

### Conclusion of Ashotsk area

According to the data we explored Ashotsk is not hot and dry place, and fire cause probability is not high. The average temperature in Ashotsk in June-July is about 15.5 C, and the average humidity is about 56%, which is considered normal. 

Wind speed is not that high, and it usually blows from West, so the fire probably spreads to East.
